Model Checking

### 2023-04-25 owen/Write Ups/01_eda.qmd
trying out some new things, fitting some new models

---
title: "Chess EDA"
author: Owen G. Ward
format: 
  # html:
  #   toc: true
  #   toc-location: left
  pdf
bibliography: refs.bib
---

@@

library(tidyverse)
library(RcppRoll)
library(cmdstanr)
library(bayesplot)

@@

- $\alpha_j$ is a player level random effect
- $\beta_j$ is a player level random effect, accounting for the win ratio $x_{ij}$
- $\gamma$ is the coefficient of game level coefficients

@@

$$
\alpha_j \sim \mathcal{N}(\mu_2, \tau_2)
$$
$$
\mu_1 \sim \mathcal{N}(0, 5)
$$
$$
\tau_1 \sim \mathcal{Cauchy}(0, 5)
$$

@@

$$
\beta_j \sim \mathcal{N}(\mu_1, \tau_1)
$$
$$
\mu_2 \sim \mathcal{N}(0, 5)
$$
$$
\tau_2 \sim \mathcal{Cauchy}(0, 5)
$$

@@

fit3_ave$summary()

mcmc_hist(fit3_ave$draws(c("alpha", "mu2", "tau2")))
mcmc_hist(fit3_ave$draws(c("beta", "mu1", "tau1", "gamma1", "gamma2")))

rhat(fit3_ave)

@@

@@

fit3_ave$summary()

mcmc_hist(fit3_ave$draws(c("beta", "mu1", "tau1", "gamma1", "gamma2")))
```

@@

fit3_ave$summary()

mcmc_hist(fit3_ave$draws(c("beta", "mu1", "tau1","gamma1", "gamma2")))
```

@@

fit3_ave$summary()

@@

>>>>

---
title: "Chess EDA"
author: Owen G. Ward
format: 
  html:
    toc: true
    toc-location: left
  # pdf
bibliography: refs.bib
---

@@

library(tidyverse)
library(RcppRoll)
library(cmdstanr)
library(bayesplot)
library(posterior)

@@

- $\alpha_j$ is a player level random effect
- $\beta_j$ is a player level random effect, accounting for the win ratio $x_{ij}$
- $\gamma$ is a fixed effect of game level coefficients

@@

$$
\alpha_j \sim \mathcal{N}(\mu_2, \tau_2)
$$
$$
\mu_2 \sim \mathcal{N}(0, 5)
$$
$$
\tau_2 \sim \mathcal{Cauchy}(0, 5)
$$

@@

$$
\beta_j \sim \mathcal{N}(\mu_1, \tau_1)
$$
$$
\mu_1 \sim \mathcal{N}(0, 5)
$$
$$
\tau_1 \sim \mathcal{Cauchy}(0, 5)
$$

@@

# fit3_ave$summary()

mcmc_hist(fit3_ave$draws(c("alpha", "mu2", "tau2")))
mcmc_hist(fit3_ave$draws(c("beta", "mu1", "tau1", "gamma1", "gamma2")))
```

The following metric (R-hat) indicates if the Bayesian model is
converging. We would like values of smaller than 1.01 here, with values
less than 1 possible in a sample. 
```{r compute r-hat}
bayesplot::rhat(fit3_ave)
```

@@

### How big are these effects

We then want to quantify how much these estimated effects change the 
expected win probability.
We have seen the size of the effects above for the non history parameters.
The largest posterior mean for a $\beta\approx 1.56$. Since the normalized win
proportion is between -0.6 and 0.6, that means that for a game
where this player has a 50% chance of winning with a normalized average win
proportion of 0, if their normalized win proportion is 0.5, the
win probability will increase by $\approx 19\%$.
This would occur, for example, if they win 50% of their total games but have
won all 10 of their last 10. This is a similar size effect to the 
increase in win probability which would come from playing an 
opponent with an ELO ranking that is 170 points lower.

Similarly, a normalized win history 0.25 
above average corresponds to a difference 
of approximately 80 in ELO scores.

The underlying parameter $\mu_1$ has posterior mean of 0.606. 
A normalised win history of 0.25 with a winner effect of this size
would increase the win probability by about 4%, a similar effect
as playing as white instead of black, all other covariates being
equal.

### Probability of Positive Winner Effects

For each of these players, we can use the posterior draws to
estimate the posterior probability that $\beta >0$, i.e
that there is a positive effect.

```{r estimate prob beta positive}

# stan_draws <- fit3_ave$draws()
# 
# 
# all_draws <- as_draws_df(stan_draws)
# 
# ndraws <- nrow(all_draws)
# 
# as_tibble(all_draws) %>% 
#   select(starts_with("beta")) %>% 
#   apply(2, function(x) sum(x>0)/ndraws)
#

prob_positive <- function(stan_draws) {
  all_draws <- as_draws_df(stan_draws)
  ndraws <- nrow(all_draws)
  as_tibble(all_draws) %>% 
  select(starts_with("beta")) %>% 
  apply(2, function(x) sum(x>0)/ndraws)
}

prob_positive(fit3_ave$draws())

We can then use this metric to compare how these winner effects change 
as we change both the length of the history and the type of games used.

@@

# fit3_ave$summary()

prob_positive(fit3_ave$draws())
```

@@

# fit3_ave$summary()

prob_positive(fit3_ave$draws())
```

@@

# fit3_ave$summary()

@@

prob_positive(fit3_ave$draws())
```

### 2023-04-25 owen/Write Ups/02_more_eda.qmd
trying out some new things, fitting some new models

---
title: "Model Checking"
author: Owen G. Ward
date: 04/17/2023
format: 
  # html:
  #   toc: true
  #   toc-location: left
  pdf
bibliography: refs.bib
---

@@

library(tidyverse)
library(RcppRoll)
library(cmdstanr)
library(bayesplot)

@@

@@

@@

Now let's fit the stan model to the permuted data and investigate what
the effects are.

@@

perm_init_data

@@

fit3_ave$summary()

mcmc_hist(fit3_ave$draws(c("beta", "mu1", "tau1", "gamma1", "gamma2")))

mcmc_hist(fit3_orig$draws(c("beta", "mu1", "tau1", "gamma1", "gamma2")))
```

So, when we permute the data in this way, get the same winner/loser effects.
This means its not a real effect. What could be causing this?

- Weird data sample?

- Not controlling for some other important predictors?

- The magnitude of these effects is the same also, which indicates its a problem

- Did I permute the data the right way then?

What about if the history is simply uniform on the [-1, 1]. What sort of
effects do you see then?

@@

@@

fit3_rand$summary()

@@

>>>>

---
title: "Model Checking"
author: Owen G. Ward
date: 04/17/2023
format: 
  html:
    toc: true
    toc-location: left
  # pdf
bibliography: refs.bib
---

@@

library(tidyverse)
library(RcppRoll)
library(cmdstanr)
library(posterior)
library(bayesplot)

@@

prob_positive <- function(stan_draws, param = "beta") {
  all_draws <- as_draws_df(stan_draws)
  ndraws <- nrow(all_draws)
  as_tibble(all_draws) %>% 
  select(starts_with(param)) %>% 
  apply(2, function(x) sum(x>0)/ndraws)
}

@@

## Identifying Problems

An issue was identified with the current implementation,
where the `previous_win_history` used, which was supposed to
be the proportion of the past $n$ games won compared to the
players overall win percentage, contained the result of the current game
(the outcome). When this is corrected the strong equal positive winner
effects are removed, and we now see a variable effect, 
with some players still displaying positive winner effects. However,
several of the players no longer show such an effect.

@@

Now we fit the stan model to the permuted data and investigate what
the effects are.

@@

# perm_init_data

@@

# fit3_ave$summary()

mcmc_hist(fit3_ave$draws(c("beta", "mu1", "tau1", "gamma1", "gamma2"))) +
  labs(title = "Permuted Data")

mcmc_hist(fit3_orig$draws(c("beta", "mu1", "tau1", "gamma1", "gamma2"))) +
  labs(title = "True Data")

```{r}
#| echo: true

## for the permuted data
prob_positive(fit3_ave)

## for the real data
prob_positive(fit3_orig)

Permuted data gives effects which are all zero on average, which is reasonable.
This would be expected as permuting the ordering of the games should
remove any possible winner effects. Similarly, the 
probability of positive effects is closer to 0.5, which
would be expected under a model with no effects.

Another way to confirm that having no history should lead to no 
winner effects is shown below.
We now replace the historic win proportion with uniform 
draws on the interval $[-1, 1]$. The estimated winner effects 
under such a history are shown below.

@@

We also see that the probability of these effects being positive is
close to 0.5, as would be expected under this model.

@@

# fit3_rand$summary()

prob_positive(fit3_rand)

@@

## Separating out history

We want to next look at the history as both the current win 
streak, up to but not including the previous game, along with the result 
of the previous game, to see if we can identify how potential winner 
effects can be decomposed between these two components.

We do this with a model of the form

$$
P(y_{ij} = 1) = \frac{1}{1 + \exp(-(\alpha_j + \beta_j x_{ij} + 
\delta_j x_{ij}^{*} + 
\gamma z_{ij}))},
$$

where now $x_{ij}^{*}$ is the result of the __previous__ game
played by player $j$, and $x_{ij}$ is their normalised 
win ratio over the previous $n$ games, not including the single
previous game.

```{r get data for winner effects}

hist_data_init <- tidy_games %>% 
  mutate(WhiteElo = as.numeric(WhiteElo), 
         BlackElo = as.numeric(BlackElo)) %>% 
  mutate(focal_user = ifelse(focal_white == 1, White, Black)) %>% 
  mutate(elo_diff = ifelse(focal_white == 1, 
                           WhiteElo - BlackElo, BlackElo - WhiteElo)) %>% 
  mutate(focal_id = match(focal_user, top_players)) %>% 
  select(focal_user, focal_id, focal_white, 
         focal_win_prop, elo_diff, focal_result) %>% 
  group_by(focal_id) %>% 
  mutate(ave_prop = lag(focal_win_prop, n = 2, default = 0) - 
           mean(focal_win_prop),
         prev_game = lag(focal_result, default = 0)) %>% 
  filter(focal_result != 0.5)

stan_data_hist <- list(N = nrow(hist_data_init),
                  J = length(top_players),
                  y = hist_data_init$focal_result,
                  id = hist_data_init$focal_id,
                  colour = hist_data_init$focal_white,
                  elo = hist_data_init$elo_diff,
                  win_prop = hist_data_init$ave_prop,
                  prev_game = hist_data_init$prev_game)

```{r fit the stan model}
#| eval: false

stan_file <- "../model4.stan"

fit4 <- mod$sample(data = stan_data_hist,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit4$summary()

fit4$save_object(file = "../model_fits/model4.RDS")

```{r load the stan fit}

fit4 <- readRDS("../model_fits/model4.RDS")

mcmc_hist(fit4$draws(c("delta"))) +
  labs(title = "Effect of Previous Game")

mcmc_hist(fit4$draws(c("beta"))) +
  labs(title = "Effect of Previous n games, excluding most recent game")

This model seems to fit well, with no potential concerns also.

We can then again compare these estimated effects. This

```{r check whether positive effects}
#| echo: true

## the beta parameters (previous n, not last game)
prob_positive(fit4)

## the delta parameters (single game) 
prob_positive(fit4, param = "delta")
```

This seems to indicate that the previous game on its own does not have as clear
an effect, but that the streak of previous games shows stronger evidence
for a winner effect.

## Consider all bullet games

Previously we considered only 60 second bullet games. We wish to 
consider if there is evidence for this effect across all bullet games.

```{r get all bullet games}

all_bullet <- lichess_data %>% 
  filter(Event == "Rated Bullet game")

tidy_games <- map_dfr(top_players, get_hist, all_bullet, prev_n = 10) %>% 
  as_tibble()

tidy_games

stan_data <- tidy_games %>% 
  mutate(WhiteElo = as.numeric(WhiteElo), 
         BlackElo = as.numeric(BlackElo)) %>% 
  mutate(focal_user = ifelse(focal_white == 1, White, Black)) %>% 
  mutate(elo_diff = ifelse(focal_white == 1, 
                           WhiteElo - BlackElo, BlackElo - WhiteElo)) %>% 
  mutate(focal_id = match(focal_user, top_players)) %>% 
  select(focal_user, focal_id, focal_white, 
         focal_win_prop, elo_diff, focal_result) %>% 
  group_by(focal_id) %>% 
  mutate(ave_prop = lag(focal_win_prop, default = 0) - mean(focal_win_prop)) %>% 
  filter(focal_result != 0.5)

```{r refit model 3 for all bullet}
#| eval: false

stan_data_ave <- list(N = nrow(stan_data),
                  J = length(top_players),
                  y = stan_data$focal_result,
                  id = stan_data$focal_id,
                  colour = stan_data$focal_white,
                  elo = stan_data$elo_diff,
                  win_prop = stan_data$ave_prop)

fit3_bullet <- mod$sample(data = stan_data_ave,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit3_bullet$save_object(file = "../model_fits/model3_n10_all_bullet.RDS")
```

```{r}
fit3_bullet <- readRDS("../model_fits/model3_n10_all_bullet.RDS")

mcmc_hist(fit3_bullet$draws(c("beta", "mu1", "tau1", "gamma1", "gamma2")))

prob_positive(fit3_bullet)
```

How do these estimates compare to those with only the 60 second games?
The model with all bullet games has all posterior means greater than 0.
There is not much of an overall change between the estimated mean
effect sizes between the two datasets.
Using all bullet games also leads to 
smaller posterior standard deviations for the
winner effect size for all players.

```{r refit model 4 for all bullet games}
#| eval: false

hist_data_init <- tidy_games %>% 
  mutate(WhiteElo = as.numeric(WhiteElo), 
         BlackElo = as.numeric(BlackElo)) %>% 
  mutate(focal_user = ifelse(focal_white == 1, White, Black)) %>% 
  mutate(elo_diff = ifelse(focal_white == 1, 
                           WhiteElo - BlackElo, BlackElo - WhiteElo)) %>% 
  mutate(focal_id = match(focal_user, top_players)) %>% 
  select(focal_user, focal_id, focal_white, 
         focal_win_prop, elo_diff, focal_result) %>% 
  group_by(focal_id) %>% 
  mutate(ave_prop = lag(focal_win_prop, n = 2, default = 0) - 
           mean(focal_win_prop),
         prev_game = lag(focal_result, default = 0)) %>% 
  filter(focal_result != 0.5)

stan_data_hist <- list(N = nrow(hist_data_init),
                  J = length(top_players),
                  y = hist_data_init$focal_result,
                  id = hist_data_init$focal_id,
                  colour = hist_data_init$focal_white,
                  elo = hist_data_init$elo_diff,
                  win_prop = hist_data_init$ave_prop,
                  prev_game = hist_data_init$prev_game)

stan_file <- "../model4.stan"

fit4 <- mod$sample(data = stan_data_hist,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit4$summary()

fit4$save_object(file = "../model_fits/model4_n10_all_bullet.RDS")

```{r examine this model fit}
fit4 <- readRDS("../model_fits/model4_n10_all_bullet.RDS")

mcmc_hist(fit4$draws(variables = c("beta", "mu1", "tau1", "gamma1", "gamma2")))

mcmc_hist(fit4$draws(variables = c("delta", "mu3", "tau3")))

prob_positive(fit4)
prob_positive(fit4, param = "delta")
```

## Just previous game

We could also only use the previous game as a predictor, 
without using the rest of the history.

```{r}
#| eval: false

stan_data_hist <- list(N = nrow(hist_data_init),
                  J = length(top_players),
                  y = hist_data_init$focal_result,
                  id = hist_data_init$focal_id,
                  colour = hist_data_init$focal_white,
                  elo = hist_data_init$elo_diff,
                  prev_game = hist_data_init$prev_game)

stan_file <- "../model5.stan"

fit5 <- mod$sample(data = stan_data_hist,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit5$save_object(file = "../model_fits/model5_n10_all_blitz.RDS")

```{r}
fit5 <- readRDS("../model_fits/model5_n10_all_blitz.RDS")

mcmc_hist(fit5$draws(variables = c("delta")))

prob_positive(fit5, param = "delta")

Fitting using just the previous game as a predictor seems to give a reasonable
model, with a similar interpretation as when we include the history also.

For this model, a posterior mean value for delta of 0.25 corresponds to 7%
increase in the win probability having won the previous game compared 
to having lost it.

Not clear if this difference is due to excluding the win
history or due to using a larger/different dataset.

### Next Steps

- Can we compare the performance of these models, using loo-cv 
or similar?

- Could this model be parameterized better?

- Any other possible problems with this?

- What about some sort of mixture of the previous game and the others before 
that?

 - Some sort of decaying effect?
